\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{algorithmic}
\usepackage{hyperref}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\usepackage{amssymb}
\newcommand*{\field}[1]{\mathbb{#1}}
\begin{document}
\title{\vspace{-5cm} Theoretical Computer Science Coursework}
\author{Jake Mortimer}
\maketitle
\begin{enumerate}
\section{Complexity and Approximability} 
\item \textbf{Question 1}
\begin{enumerate}
\item 
i. Prove that SHORT PATH $\in \textbf{NL}$ \\
Given an instance of the SHORT PATH we need to show that there is a non-deterministic Turing Machine that decides it in log-space. Starting at the start vertex $s$ of G we have a counter $c$ in memory. At each vertex in the path from $s$ we non-deterministically guess the next vertex. If we reach $t$ and the value in the counter $c \leq k$ then accept. If $c>k$ then reject. This obviously solves SHORT PATH in logarithmic space because the counter can have a maximum value of $|V|$. If the counter were greater than $|V|$ then this would mean that we have visited one vertex of the input graph more than once and, hence, there is a cycle. $\therefore $ We could reduce the path length to some path length $\leq |V|$ by simply removing the cycle $\Rightarrow c$ can be stored in logarithmic space and SHORT PATH $\in \textbf{NL}$.\\
ii. Describe the reduction REACHABILITY $\leq _{L} $ SHORT PATH \\
We are initially given an instance of REACHABILITY; this is a graph $G = (V,E)$ and two vertices $s,t$. A yes instance of REACHABILITY is where $t$ is reachable from $s$ by a directed path on the edges of $G$. To create an instance of SHORT PATH in logarithmic space we use the following reduction:\\
The instance of SHORT PATH consists of the graph $G'=(V',E')$, the vertices $s',t' \in V'$ between which we find a path of length at most $k=|V|+1$. For each vertex $v \in V$ we create the vertices $v^1,v^2,...,v^{|V|}$ and add these to $V'$. Also, create the vertices $s',t'$ and add them to $V'$. For each directed edge $(u,v) \in E$ we create the undirected edges $(u^i,v^{i+1})$ for all $i \leq |V|-1$ and add these to $E'$. Also, create the edges $(s',s^1),(t^1,t^2),(t^2,t^3),...,(t^{|V|},t')$ and add these to $E'$. Finally, to complete the reduction we set $k=|V|+1$.
iii. Prove the correctness of the above reduction\\
In order to prove the correctness of the above reduction I will prove that a yes instance of REACHABILITY $\Rightarrow $ yes instance of SHORT PATH, and a no instance of REACHABILITY $\Rightarrow $ no instance of SHORT PATH. This is sufficient to show that SHORT PATH is $\textbf{NL}$-hard, and, when combined with the proof of membership of $\textbf{NL}$ above, proves that SHORT PATH is $\textbf{NL}$-complete.\\
\textbf{Note}: $v^i$ means the vertex $v$ with \textit{row} $i$

First we prove that a yes instance of REACHABILITY $\Rightarrow $ yes instance of SHORT PATH:\\
Suppose that we have a yes instance of REACHABILITY. This would mean that $t$ is reachable from $s$ in the directed graph $G$ (defined in the instance of REACHABILITY). Let us construct an instance of SHORT PATH using the above reduction. This would yield the instance $G'=(V',E')$, $k=|V|+1$, and the start and end vertices $s',t'$. If $t$ is reachable from $s$ in $G$ then this means that there is some path of directed edges starting at $s$ and ending at $t$. This directed path $P=(s,u),(u,v),...,(w,t)$ for some $u,v,w \in V$ would become the following path in $G'$ $P'=(s',s^1),(s^1,u^2),(u^2,v^3),...,(w^{l-1},t^l),(t^l,t^{l+1}),...,(t^{k-1},t^k),(t^k,t')$. As can be seen the superscript, which corresponds to the rows of the vertices, monotonically increases by 1 each time if the edges traversed in $G'$ are traversed in the correct direction as defined by the digraph $G$. If some row were visited more than once this would mean that one of the edges corresponding to the directed edge in $G$ would have been traversed in the wrong direction. For the constructed instance to be a yes instance $|P'|\leq k =|V|+1$. If the latter of the above cases were true and we visited one row multiple times then this would add at least 2 to the length of $P'$. There are $|V|+1$ rows in the graph excluding the start vertex; if we were to traverse an edge in the wrong direction we would get $|P'| \geq |V|+3 > |V|+1 \Rightarrow$ this would be a no instance of SHORT PATH. $\therefore$ we can only have a yes instance if all the edges of $G'$ are traversed so the row value increases, which corresponds to the correct direction in $G$. \\
Secondly, we need to prove that a no instance of REACHABILITY $\Rightarrow$ no instance of SHORT PATH: \\
Suppose that we have a no instance of REACHABILITY. This would mean that $t$ is not reachable from $s$, which in turn means that there is no path from $s$ to $t$ in $G$. Let us construct the corresponding instance of SHORT PATH using the above reduction. This would yield the instance $G'=(V',E')$, $k=|V|+1$, and the start and end vertices $s',t'$. \\
Assume, for a contradiction that there is a path in $G'$ from $s'$ to $t'$ of length at most $k=|V|+1$, which we call $P'$. This would correspond to a yes instance of SHORT PATH. As such, using the logic above each row can only be visited once otherwise $|P'|>k=|V|+1$, which would be a no instance of SHORT PATH. We reverse the reduction from SHORT PATH to REACHABILITY to get $G,s,t$, and transform the path $P'$ to $P$ on $G$, in the following way $P'=(s',s^1),(s^1,u^2),(u^2,v^3),...,(w^{l-1},t^l),(t^l,t^{l+1}),...,(t^{k-1},t^k),(t^k,t')$ would become This directed path $P=(s,u),(u,v),...,(w,t)$ for some $u,v,w \in V$. This would trivially yield a path from $s$ to $t$ in $G$, and hence we would have a yes instance of REACHABILITY. This is a contradiction $\Rightarrow$  no path from $s'$ to $t'$ in at most $k$ steps in the graph $G'$. $\therefore$ the constructed instance of SHORT PATH is a no instance.\\
This means YES $\Rightarrow$ YES and NO $\Rightarrow$ NO, and so we have correspondence of yes instances of these problems. \\$\therefore$ SHORT PATH is $\textbf{NL}$-complete $\square $

\item Show that SHORT-PATH-2019 $\in \textbf{L}$ \\
The following proof that SHORT-PATH-2019 $\in  \textbf{L} $ will be constructive, as such I will present an algorithm to solve the problem that works in logarithmic space.\\
I will first introduce a useful syntactic structure:\\
$for$ $v_1,v_2,...v_l \in V$ $\equiv $ $for$ $v_1 \in V \{ for$ $v_2 \in V \{ ... \{for$ $v_l \in V \{...\}\}\}\} $\\
The algorithm is as follows:
\begin{algorithmic}
\FOR {$v_1,v_2 \in V$}
	\IF {sequence of vertices is valid path from $s\rightarrow t $}
		\STATE Accept
	\ENDIF
\ENDFOR
\FOR {$v_1,v_2,v_3 \in V$}
	\IF {sequence of vertices is valid path from $s\rightarrow t $}
		\STATE Accept
	\ENDIF
\ENDFOR
...
\FOR {$v_1,v_2,v_3,...,v_{2017} \in V$}
	\IF {sequence of vertices is valid path from $s\rightarrow t $}
		\STATE Accept
	\ENDIF
\ENDFOR
\FOR {$v_1,v_2,v_3,...,v_{2017},v_{2018} \in V$}
	\IF {sequence of vertices is valid path from $s\rightarrow t $}
		\STATE Accept
	\ENDIF
\ENDFOR
\STATE Reject
\end{algorithmic}
The sequence of vertices that are checked when determining the validity of a path are $(s,v_1,...,v_l,t)$. These paths are of length at most 2019 $\Rightarrow $ the space used to store each enumerated path is bounded by a constant $\Rightarrow $ the construction trivially uses logarithmic space and solves the problem SHORT-PATH-2019.
$\therefore $ SHORT-PATH-2019 $\in \textbf{L}$ $\square$

\end{enumerate}
\item \textbf{Question 2}
\begin{enumerate}
\item Prove unconditionally that TWO CLIQUES is not \textbf{PSPACE}-complete with respect to log space reductions.\par
We can show that TWO CLIQUES problem is a member of $\textbf{L} $. This can be shown by the proving the reduction TWO CLIQUES $\leq _{L}$ BIPARTITE which is a member of $\textbf{L} $. BIPARTITE is defined as follows:\\
\textit{Instance:} Graph $G=(V,E)$\\
\textit{Question:} Is the graph G bipartite?\\
The reduction is simple; given an instance of TWO CLIQUES (an undirected graph $G$) we take the complement of the graph, resulting in graph $\bar{G}$ which we give as the instance of BIPARTITE.\\
\textit{\textbf{Proof that BIPARTITE $\in \textbf{L} $}}\\
We will be using Reingold's Theorem that $s-t$ connectivity in an undirected graph is in $\textbf{L} $, \href{https://en.wikipedia.org/wiki/SL_(complexity)}{seen here - \textit{SL and Reingold}}. With the knowledge that UPATH (does there exist a path from $s-t$ in $G$?), we prove that BIPARTITE $\in \textbf{L} $ by reducing to co-UPATH (The complement of UPATH; there is no path from $s-t$ in $G'$), as $\textbf{L} = co-\textbf{L} $ trivially. \\
For a given bipartite graph $G=(V,E)$ we can define the graph $G'$ such that $V'=\{v_1,v_2\mid v \in V\} \cup \{s,t\} $ and $E'=\{(u_1,w_2),(u_2,w_1) \mid (u,w)\in E\} \cup \{(s,v_1) \mid \forall v \in V\} \cup \{(v_2,t) \mid \forall v \in V\}$.
This reduction can clearly be done in logarithmic space. Now we have a reduction we need to prove the correspondence of yes-instances: \\
($\Rightarrow $) Given a yes instance of BIPARTITE, a graph $G$, we can use the reduction above to get the graph $G'$. We substitute this graph into the co-UPATH problem. We pass in the vertices $s,t$, and if it accepts; there are no paths between any of the expanded vertices. This is the case because for there to be a path between $s$ and $t$ there would need to be a path between two vertices such as $v_1,v_2 \in V'$ there must be an edge within one of the sets of vertices of $G$. \\Assume, for a contradiction, that there was not an edge between any two vertices of the same set in $G$ but there was a path $v_1\rightarrow v_2$. The same disjoint sets in $G$ can be applied to $G'$, and we have established that there are no edges within a set in $G$, similarly, this is a fact for these sets in $G'$. This means that $v_1,v_2$ would have to be connected by some vertex in the opposite set $(v_1,d_1),(v_2,d_1)$ for there to be a path between them. This is not possible: it would mean that in the original graph $G$ there would have been the edges $(v,d),(d,v)\in E$, and in an undirected graph there would only be one of these edges. Therefore, we have a contradiction meaning that for there to be a path $v_1\rightarrow v_2$ there would have to be an edge within the bipartite sets in $G$, and in turn a yes instance of BIPARTITE $\Rightarrow $ yes instance of co-UPATH. \\
($\Leftarrow $) Given a yes instance of co-UPATH which corresponds to the above reduction we transform $G'$ to $G$. For the instance to be a yes instance, there can be no path $s\rightarrow t$ which means no pair of vertices $v_1,v_2 \in V'$ can have a path between them, otherwise there would be a path from $s-t$. When we reduce $G'$ to $G$ we first remove $s,t$, this leaves two sets of disjoint vertices with no edges between vertices of the same set. Assume, for a contradiction, that the two sets had an edge within one of the sets, this would mean that $(v_1,w_2)$ would be an edge for some $v_1,w_2 \in V'$ which are in the same set. From the construction, this would mean that there is a path from $s-t$. $\therefore $ we have a contradiction $\Rightarrow $ there can be no edges within these disjoint sets. We then construct $G=(V,E) $ from the pairs of vertices $v_1,v_2 \in V'$, and the edges $(v_1,w_2),(v_2,w_1) \in E'$ as described in the reduction. This results in a graph $G$ which, trivially, is bipartite: two disjoint sets maintained from $G'$ and there can be no edges between two vertices of a set. $\therefore $ yes instance of co-UPATH $\Rightarrow $ yes instance of BIPARTITE. \\
$\therefore $ BIPARTITE $\leq_{L} $ co-UPATH $\leq_{L} $ UPATH $\Rightarrow $ BIPARTITE $\in \textbf{L} $ $\square $\\
\textit{\textbf{Continuing from before proof...}}\\
We now prove the reduction TWO CLIQUES $\leq _{L}$ BIPARTITE. \\ 
\textit{\textbf{Proof of Reduction}}\\  
To prove the correctness of this reduction we need to prove the correspondence of yes instances. This means that we have a yes instance of TWO CLIQUES $\iff $ we have a yes instance of BIPARTITE.\\
($\Rightarrow $) Given a yes instance of TWO CLIQUES $G, V_1, V_2$ where $G$ is the graph, and $V_1,V_2$ are the two cliques that the graph can be subdivided into. We take the complement of $G$, $\bar{G}$ as in the reduction. In the initial yes-instance for TWO CLIQUES $V_1$ and $V_2$ are  completely connected sub-graphs; each vertex of either is connected to every other vertex of the same clique. When we take the complement of $G$, the edge set contains the edges which were not present in the initial graph. This means that $V_1,V_2$ will have no edges internally in $\bar{G}$ and only edges between vertices of opposite sets. As such the graph $\bar{G}$ is bipartite. $\therefore $ yes instance of TWO CLIQUES $\Rightarrow $  yes instance of BIPARTITE.
\\
($\Leftarrow $)Given a yes instance of BIPARTITE $\bar{G}, V_1, V_2 $ we take the complement of the graph $\bar{G} $ and get $G$. By the reverse of the above argument it is clear that a yes instance of BIPARTITE $\Rightarrow $ yes instance of TWO CLIQUES.
\\
$\therefore $ TWO CLIQUES $\in \textbf{L} $ $\square $ \\
\textit{\textbf{Continuing from before proof...}}\\  
We now have that TWO CLIQUES $\in \textbf{L}$. It is known that $\textbf{L} \subseteq \textbf{NL} \subseteq \textbf{PSPACE} $ because if a problem is solvable in logarithmic space on a deterministic machine it is obviously solvable in logarithmic space on a \textit{non}-deterministic machine $\Rightarrow $ $\textbf{L} \subseteq \textbf{NL} $. It is also clear to see that if a problem is solvable in logarithmic space on a non-deterministic machine then it is solvable in polynomial space on a deterministic machine by Savitch's Theorem $\Rightarrow$  $\textbf{NL} \subseteq \textbf{PSPACE} $. \\
\\
If $\textbf{L} \subseteq \textbf{NL}$ and $\textbf{NL} \subseteq \textbf{PSPACE}$, then the above sequence of set containments holds true.\\
We now attempt to derive a contradiction: assume that TWO CLIQUES is \textbf{PSPACE}-complete. This would mean that $\textbf{L} = \textbf{PSPACE} \Rightarrow \textbf{PSPACE} \subseteq \textbf{NL}$. We already have that $\textbf{NL} \subseteq \textbf{PSPACE}$, so combining these we get that $\textbf{NL} = \textbf{PSPACE} $ which we can prove not to be true.\\
\textit{\textbf{Proof of $\textbf{NL} \neq \textbf{PSPACE}$}}\\
First, we need to define space constructible functions. A function $f:\field{N} \rightarrow \field{N} $ such that $f(n) \geq log(n)$ is called space constructible if the function that maps $1^n$ to the binary representation of $f(n)$ is computable in space $O(f(n))$ From this it is clear that if $f(n)$ and $g(n)$ are space constructible functions, and $f(n)=o(g(n))$ then $SPACE[f] \subset SPACE[g]$. By Savitch's Theorem, $\textbf{NL} \subseteq SPACE[log^2(n)]$. Since $log^2(n) = o(n)$, we have $SPACE[log^2(n)] \subset SPACE[n]$ $\therefore \textbf{NL} \neq \textbf{PSPACE}$ $\square $  \\
\textit{\textbf{Continuing from before proof...}}\\
This is a contradiction $\therefore $ we can say (unconditionally) that TWO CLIQUES is not $\textbf{PSPACE}$-complete $\square $
\end{enumerate}
\section{Algorithmic Game Theory} 
\item \textbf{Question 3}
\begin{enumerate}
\item Find all the pure strategy equilibria. \\
There are $n \choose 2 $ pure strategy equilibria in this game; where exactly two people are helping. This is the case because in these circumstances if one of the players were to stop helping then they would decrease their score from 9 to 8 $\Rightarrow $ this would no longer be an equilibrium. Also, if one of the players who was currently not volunteering changed to volunteering then they would decrease their score from 10 to 9 $\Rightarrow $ this would no longer be an equilibrium.
\item Find all the symmetric mixed strategy equilibria. \\
\textit{\textbf{Note:}} $P(n)$ is the probability function describing the probability of $n$ players volunteering. \par
\textit{Score for volunteering}: 9\\
\textit{Score for not volunteering}:\\ $10(P(\geq 2)) + 8(P(<2)) = 10(P(\geq 2)) + 8(1-P(\geq 2))$ \\
$10P(\geq 2) + 8(1-P(\geq 2)) = 9$\\
$8(1-P(0)-P(1))+10(P(0)+P(1)) = 9$\\
$10P(0)-8P(0)+10P(1)-8P(1)+8 = 9$\\
$2(P(0)+P(1))=1$\\
$P(0)+P(1) = \frac{1}{2}$\par
Now using the binomial distribution (${n \choose r}q^r (1-q)^{n-r}$) we can see that the probability of $P(0)$ and $P(1)$ are as follows:\\
$P(0) = p^{n-1}$, $P(1)= (n-1)(1-p)p^{n-2} $\par
Using these we can create the following polynomial to attempt to solve for $p$ which is the probability of \textbf{not} volunteering:\\
$p^{n-1}+(n-1)(1-p)p^{n-2} -\frac{1}{2} = 0 $\\
$(4-2n)p^{n-1} + 2(n-1)p^{n-2} - 1 = 0$\\
$2(n-2)p^{n-1}+2(1-n)p^{n-2} +1 = 0$\par
\textbf{Intermediate Value Theorem} can now be used to show that a solution to this polynomial exists:\\
$F(p)=2(n-2)p^{n-1}+2(1-n)p^{n-2} +1$\\
$F(0)=1$\\
$F(1)=2(n-2)+2(1-n)+1 =-4+2+1=-1$\\
$\therefore $ $y=F(p)$ is such that $F(1) < F(p) < F(0)$ meaning there is some $c$ between $0<c<1$ for which $c$ is a solution of $F(p)$ and since $F(p)=0$ above must be solution to $F(p)=0$ for our symmetric mixed strategy; our probability $\square$
\end{enumerate}
\item \textbf{Question 4} \\
Show that a Boolean function $f(p_1, p_2)$ can be computed by a three-player game. That is, each player has two pure strategies that are the Boolean values $T$ and $F$, and there’s a pure-strategy Nash equilibrium if and only if the strategy of player three, $p_3 =f(p_1, p_2)$ where $p_1$ and $p_2$ are the strategies of players one and two, respectively.\par

The construction of the game is as follows: \\
We have three players 1,2,and 3. Each player has two pure strategies $T$ and $F$. The players receive a score of 1 if the strategies $p_1,p_2,p_3$ are such that $p_3=f(p_1,p_2)$, and a score of 0 otherwise.\\
Using the \textbf{above} construction we want to prove that there is a pure strategy Nash Equilibrium $\iff $ the strategy of player 3, $p_3$, is such that $p_3=f(p_1,p_2)$.\\
($\Rightarrow $)\\
Pure Strategy Nash Equilibrium (NE) $\Rightarrow $ $p_3=f(p_1,p_2)$\\
Suppose that there exists some pure strategy NE such that $p_3 \neq f(p_1,p_2)$. This would mean that the score for the players would be 0 as such there would be a unilateral change in strategy that results in a higher score $\Rightarrow $ this is not a pure strategy NE. $\therefore $ we have a contradiction which means that Pure Strategy NE $\Rightarrow $ $p_3=f(p_1,p_2)$.\\
($\Leftarrow $)\\
$p_3=f(p_1,p_2)$ $\Rightarrow $ Pure Strategy Nash Equilibrium\\
If $p_3=f(p_1,p_2)$ then the players would get a score of 1 which is the maximum score in the game, this means that any change in strategy would result in a lower or equal score $\Rightarrow $ this is a pure strategy equilibrium. $\therefore $ $p_3=f(p_1,p_2)$ $\Rightarrow $ Pure Strategy NE.\\
As demonstrated we have proven the correspondence of $p_3 =f(p_1, p_2)$ and a pure strategy NE, and in turn we have shown that a Boolean function $f(p_1,p_2)$ can be computed by a three-player game $\square $
\end{enumerate}
\end{document}

